CMSC818s: Project 3 \\ A Brief Survey of Modern Virtualization Research

### 2011-12-13 paper.tex
greg's edits

The concept of virtualization has been around for a long time, reaching back
to the 1960s and the introduction of virtual memory in timeshared machines.
However, in the last decade or so it's become a very exciting area of research,
particularly motivated by a new market for server virtualization.  Commerical
products like VMWare and open-source platforms like Xen \cite{ref:xen} have made
it possible for companies to host dozens of instances of their servers
on the same hardware.  Companies can now spend less on metal, can seamlessly and
easily migrate servers from one machine to another, and can even vary their fleet
size on the fly in response to fluctuations in server traffic by spinning up new
instances.  In addition to the commercial successes that have come as a result
of server virtualization, there have been tremendous breakthroughs in the research
community as well, ranging from new solutions to problems of isolation and protection
to novel new algorithms for scheduling and resource allocation.

Despite the flurry of interest in this area, there are still many interesting
and difficult problems to be solved.  Sharing finite resources
fairly and safely amongst many competing processes has never been easy, but
the popularity of virtualization in the cloud makes scalability and efficiency
bigger concerns than ever before.  Furthermore, the commercialization of
virtual machines raises questions of how to ensure quality of service in
an environment where everyone is in contention for the hardware.  In this
paper, we will examine some of the recent literature in this area and
seek to draw some conclusions about the problem space and the directions
of future work.

The rest of the paper is structured as follows: in Section \ref{sec:background}
we provide background information on the history of virtualization and the
sorts of techniques which are utilized today.  Section \ref{sec:summaries}
presents summaries of a foundation virtualization paper from the
5\textsuperscript{th} USENIX Symposium on Operating Systems Design
and Implementation, as well as three recent papers from the virtualization
session of the same Symposium in 2010.  Section \ref{sec:relns} attempts
to analyze and classify these papers in terms of how they
relate to each other and how they approach the larger problem space.
Finally, in Section \ref{sec:conclusion} we conclude by discussing our
thoughts and opinions on the problem space as a whole, and postulate
about how virtualization may evolve in the future.

@@

The concept of virtualization reaches back to the 1960s, when developers
of large time-sharing systems sought ways to give each user the illusion that
they had sole access to a private, virtual computer, despite the fact that
many users were in fact sharing computing resources.  Once private physical
computers became available, popular, and cheap in the 80s, virtualization
took a backseat to other areas of research.  However, many of the ideas
put forth in these early days lived on in work involving virtual memory
and operating system protection paradigms.

Server virtualization arrived on the scene about a decade ago, with VMWare
appearing in 1999 and Xen in 2003.  Since then, research and commercial interest
in virtualization has sparked significantly, especially in conjunction
with the rise of cloud computing and infrastructure-as-a-service.

In a virtualized context, we have many \emph{virtual machines}(VM) sharing the same
physical hardware.  In a \emph{fully virtualized} system, each VM has the illusion
that it alone has sole access to the hardware, although in a \emph{paravirtualized}
VMs may be modified to be aware of the fact that they are virtualized.  This
sharing of resources is possible because access to the hardware is regulated by
a special piece of software, the \emph{virtual machine monitor} or \emph{hypervisor}.
In many ways, the hypervisor plays the role of the kernel in a traditional operating
system - it contains the privileged software that can directly access the hardware.

The hypervisor (and potentially other administrative software) runs in the \emph{host}
domain (in Xen, this is called \emph{dom0}), while other OSes running in VMs are 
referred to as \emph{guests} (in Xen, all guests run in \emph{domU}).  Generally, it is
crucial that all guests are isolated from one another, so that the actions of one guest
cannot adversely affect the performance of any other.  This is a difficult property
to provide in a system where all guests are in contention for hardware resources.

In the early days of virtualization, there was little-to-no hardware support,
and so the multiplexing of VMs onto the physical hardware was accomplished by
\emph{emulating} hardware devices in the software for each VM.  This had the unfortunate
drawback of being extremely slow and inefficient, because the hardware was emulated by
the hypervisor, and so most device accesses by a guest VM resulted in multiple traps
into the hypervisor.  In the past decade or so, chipsets have become more friendly,
and most architectures (notably the x86) now offer hardware-level support for
virtualization, along with a set of privileged instructions for making use of it.
This has allowed hardware virtualization to become fast enough to be viable,
so that virtualized systems are now reasonably competitive with their native
counterparts.

@@

@@

Turtles is designed for the x86 architecture, which subscribes to a
single-level hardware support model.  This means that there are only two
levels of execution privilege.  Turtles runs its host hypervisor, which
we will call $L_0$, at the most privileged level of execution.  Now, suppose
we have a hypervisor running above $L_0$ at level $L_1$, and that hypervisor
is running a guest OS at level $L_2$.  Everything running at $L_1$ and $L_2$
has to run at the same privilege level, and only $L_0$ is privileged enough
to interact with the hardware.  Essentially, levels $L_1$ through $L_n$ only
exist logically, and in reality they are all multiplexed to be running directly
on top of $L_0$ (see Figure \ref{fig:turtles-multiplex}).  All traps generated at
$L_2$ go immediately to $L_0$, which
may handle them itself, or forward them back to $L_1$ for handling if hardware
interaction is not required.

@@

There are two difficult problems here that turtles provides a unique solution for.
The first involves page tables for the MMU.  In a nested model, virtual addresses for $L_2$
will need to be translated three times (first to physical addresses at $L_2$, then to $L_1$
addresses, and finally to $L_0$ addresses)
before they are readable by the hardware.  Modern MMUs provide support for virtualization
in the form of two distinct translation tables to make this process fast, but this is not
sufficient for more than one level of virtualization.  Furthermore, providing these
translation tables by emulating them in software is slow, particularly on updates,
for the reasons described above.  The paper's novel solution here is to use
\emph{multidimensional page tables}.  Essentially, each level keeps its own
shadowed page tables in software, but they are read-only.  On a page fault in $L_2$,
$L_0$ catches the trap and forwards it to $L_1$.  If $L_1$ also faults, $L_0$ allows
it to update its shadow table with the $L_2$-to-$L_1$ translation, but then applies
its tranlation for $L_1$ addresses to update the hardware table with the correct
$L_2$-to-$L_0$ translation before returning.  In this way, the hardware table at $L_0$
contains all the shadow tables above it compressed into a single table, so that
any MMU accesses can look up the correct translations directly from the hardware
without having to go through cascading levels of shadow tables first.  Turtles
also employs ``huge pages'' and Virtual Processor ID tags on page table entries
to ensure that page tables remain small and fast.

The second difficulty deals with device I/0.  Traditional approaches to this problem
involve either software emulation (which is slow), installing modified drivers in the
guest (which is not preferrable), or assigning devices to guests as they have need
of them (which is hard in the nested case).  Turtles uses something called
\emph{multi-level device assignment}, which is essentially an adaptation of the
multidimensional page table idea to the device I/O problem.  Modern chipsets
include a component called an IOMMU, which resides between the DMA and main memory
and performs translations just like the MMU does for the CPU.  By shadowing read-only
IOMMU translation tables at higher levels, and then compressing them into a single
hardware-level table, the DMA can avoid the same cascading translation issue that
would affect the MMU, thus making DMA I/O access very efficient.  For Memory-Mapped I/O
and Network Port I/O, address translation is not an issue, and so here techniques for
single-level device assignment are used without any penalty.

The Turtles paper provides an extensive evaluation section, with thorough experimentation
comparing performance of applications running on the native OS, a single-level VM, and the
Turtles two-level nested VM.  The authors measure and discuss performance on microbenchmarks,
I/O intensive workloads, and workloads that are designed to fault to the kernel constantly.
The paper also analyzes the success of multidimensional paging and multi-level device
assignment by testing these techniques against more traditional ways of handling these
problems.  In general, the Turtles hypervisor far outperforms any other approach to
nested virtualization considered, and is able to minimize the overhead of a second
level of virtualization to as low as 6-8\%, which is rather impressive.  No experiments
were conducted involving more than two nested levels of virtualization, although analysis
seemed to indicate that most of the additional overhead incurred by adding a second level
came from cache pollution and the overuse of privileged instructions where it was not
necessary.

@@

The mClock paper \cite{ref:mclock} is not targeted at nested virtualization,
but further explores
the issue of fair I/O device assignment raised in the Turtles paper.  This turns
out to be a very difficult problem even in the single-level case, particularly
when applied to virtualization in the cloud, where networked devices are
often shared not just between VMs running on one host, but between many hosts
across the network.  In the interest of fairness, a good hypervisor will try to
share device throughput equally across its guests, but in the networked case, the
total amount of capacity to share may fluctuate due to traffic from other
networked clients.  Furthermore, many OSes run their own device access algorithms
above the hypervisor level to improve performance; for example, physical disk
accesses may be reordered by the kernel to take advantage of spatial locality
when moving the disk head.  These sorts of optimizations are often difficult
for the hypervisor to multiplex without loss of efficiency.

The mClock algorithm seeks to fairly schedule the device I/O of multiple guests,
while maintaining each guest's limit (maximum throughput requirement), reservation
(minimum requirement), and weight (proportional share of capacity relative to
other guests).  The algorithm uses a ``novel, lightweight'' tagging scheme to
achieve this despite fluctuation in total capacity.  Previous attempts to solve
this problem were not able to achieve both of these goals simultaneously.

@@

The mClock algorithm (see Figure \ref{fig:mclock-alg}) works by assigning numerical tags to each
queued request.
Essentially, tags for VM $i$ with proportional weight $w_i$ will be spaced by $1/w_i$.
Requests will be executed in the order of their assigned tags, which ensures that
VM $i$ receieves his proportional share.  This approach requires a global clock
for tag assignment, so that idle clients remain synced with active ones and do not
amass ``idle credit'', thus gaining preferential treatment.  The tag for VM $i$ and
request $j$ is set as follows:

\[
	Tag(i, j) = max(curr~time, Tag(i, j-1) + 1/w_i)
\]

After the tagging, the request with the lowest tag will be scheduled.

@@

The paper describes some further tweaks to the algorithm, such as allowing
a limited amount of idle credit to be saved up by idle VMs, or breaking ``large''
requests into a series of smaller requests to account for request latency
in the scheduler.  There's also a modifed dmClock algorithm for use with
a distributed storage system, which multiplexes VM requests fairly across multiple
storate servers.  Finally, there's a heuristic for setting reservations for
all VMs by default, so that no VM will be starved.

@@

The algorithm is successful at delivering on its guarantees despite contention,
which is important for ensuring VM isolation, and also has implications in the cloud,
where meeting quality of service guarantees determine financial success.

@@

The ``Virtualize Everything but Time'' paper \cite{ref:time} notes that accurate
timekeeping has become very important in today's computing systems, with
applications to everything from systems measurement and distributed database
consistency to high-speed trading and billing for cloud services.  However, 
timekeeping in a VM context is very difficult.  Historically, timekeeping
is accomplished using software clocks based on hardware oscillators, and
keeping the software clocks synchronized to an online
reference clock.  However, the synchronization algorithms used rely on
computing delays between timestamped packets, and the delay computation
turns out to be extremely unstable in a virtualized context, due mostly
to added latency during packet processing.

@@

The focus of the paper is more on timekeeping synchronization algorithms.
The most popular contender in this arena is the \emph{Network Time Protocol} (NTP).
NTP is a feedback algorithm, which works by collecting and timestamping a series
of packets from a networked reference clock, computing clock error using
the reference timestamp-local timestamp pairings as data points, 
and updating the system clock accordingly.  Xen Clocksource makes use of NTP,
although it must be adapted to fit the virtualized context.  There are two
possibilities for this:  in Dependent Clocksource, \emph{dom0} runs an NTP
clock, and all guests running in \emph{domU} sync periodically to this NTP
clock and use Clocksource to interpolate between syncs.  Alternatively, there
is an Independent Clocksource paradigm in which each \emph{domU} guest runs their own
NTP clock.  The paper conducts evaluation of both paradigms, and argues
that neither is well-suited to the virtualized context.  Dependent Clocksource
exhibits cyclical sawtoothed drift because the Clocksource algorithm in
\emph{domU} does not have direct access to the TSC, and so cannot adjust
the NTP clock data intelligently.  Independent Clocksource is extremely
inefficient, but even worse, additional latencies incurred by contention for
network I/O tend to push the Clocksource feedback algorithm into instability,
so that large error can accumulate even across NTP synchronizations.

@@

As we surveyed this series of papers, we noticed that the literature
in this area seems to neatly classify itself by \emph{motivation}.
Many papers on virtualization come from the realm of academia, and these
are mostly characterized by a focus on clean abstractions and solutions
with good theoretical properties, such as fairness and strong isolation.
Many more papers, however, originate from corporate research at places
like Microsoft and HP Labs.  These papers seem much more interested
in providing fast, scalable solutions using commodity hardware and
software.  In our papers, we found a series of tensions which seemed
to be expressing this fundamental question of motivation.  We will
now turn our attention to these tensions, keeping in mind that the
Turtles and mClock papers share corporate authorship, while the
Denali and Time papers have originated in the realm of academia.

@@

The most obvious tension which surfaced in the literature was a question
of priority between performance and fairness.  Generally speaking, there is a trend
in corporate papers to be heavily concerned with speed and efficiency, while
academically-minded papers seem more focused on strong properties of isolation
and guarantees of fairness between VMs.  This should not be particularly surprising,
given the authorship; more efficient virtualization seems to be an easily reachable
step toward greater profit.  It is worth noting, however, that in the more recent
literature corporate papers seem to be taking note of the importance of fairness;
we will discuss this further in Section \ref{sec:relns/scale}.

Our papers are clearly divided by this metric:  Turtles is all about efficiency
and performance, often at the cost of fairness.  The problem seems
especially hard here because of the nested context.  It seems perfectly logical that if
a guest OS running in $L_2$ should only receive a fraction $\alpha$ of the resources allocated
to its host in $L_1$, which in turn should only receive a percentage $\beta$ of the machine's
total resources, then these two fractions should compound and our guest should end up with
$\alpha\beta$ of the total capacity available.  However, in reality Turtles multiplexes
all guests at all levels onto $L_0$, so in many ways the guest OS and its hypervisor will
be treated as siblings rather than a parent and child.  Multi-level device assignment
for Port I/O and Memory-mapped I/O is particularly agnostic to these concerns, because
it works exactly the same way as single-level device assignment, and so doesn't care whether
a guest runs at $L_1$ or $L_{100}$.  Furthermore, no experimentation at all is done to analyze
the effects of this model on fairness, and the authors even make the statement that this
is a secondary concern, and so they are content with a sort of best-effort.

On the other hand, Denali and mClock are all about prioritizing fairness.  The foundational
goal of the Denali isolation kernel is to ensure strong isolation between clients, so that
the behavior of any particular VM cannot in any way affect another.  And the whole point of
the mClock scheduling algorithm is to ensure that limits, reservations, and weights
are met fairly between VMs contending for a resource.  It should be noted that neither
of these papers particularly eschew performance as a worthy goal; rather, the view
here seems more to be that fast performance for a particular guest doesn't really
mean much if a malicious (or even just selfish neighbor) can somehow choke that VM
out and make it impossible to make progress.

@@

The split here is pretty clear as well: Denali, Xen, and Time all consider
paravirtualization to be beneficial and even necessary in the right situations.
In Denali, the isolation kernel presents a modified and simplified hardware abstraction
to its clients, which cleverly allows for some big wins in efficiency and performance
as a side effect.  Xen claims that paravirtualization is necessary to ensure fairness
and prevent starvation in process and I/O scheduling; one could argue, however, that
this is no longer a concern due to the advent of newer virtualization-aware hardware
(see Section \ref{sec:relns/hard-emul}).  The Time paper goes a step further and claims that
even with the new hardware, making the guest OS aware of the virtualized context it
runs in is necessary to ensure the correctness of certain time-dependent applications!

In contrast, the Turtles project argues for full virtualization and unmodified guest
OSes and hypervisors.  In the paper, the argument for this goes back to the
motivating question of efficiency, though it was unclear to us why paravirtualization
would have any negative impact here.  But it seems also that for any sort of large-scale
commercial deployment of server virtualization technology, the ability to use commodity
guests would be crucial in order to make adoption easy, and we consider this to also
be a likely factor influencing the design choice.

>>>>

The idea of virtualization has been around for a long time, reaching back
to the 1960s and the introduction of virtual memory in timeshared machines.
However, in the last decade or so it's become a very exciting area of research,
particularly motivated by a new market for server virtualization.  Commerical
products like VMWare and open-source platforms like Xen \cite{ref:xen} have made
it possible for companies to host dozens of instances of their servers
on the same hardware.  Companies can now spend less on metal, can seamlessly and
easily migrate servers from one machine to another, and can even adjust their fleet
size on the fly in response to fluctuations in server traffic by spinning up new
instances.  In addition to the commercial successes that have come as a result
of server virtualization, there have been tremendous breakthroughs in the research
community as well, ranging from new solutions to problems of isolation and protection
to novel algorithms for scheduling and resource allocation.

Despite the recent flurry of activity in this area, there are still many interesting
and difficult problems to be solved.  Sharing finite resources
fairly and safely amongst many competing processes has never been easy, but
the popularity of virtualization in the Cloud makes scalability and efficiency
bigger concerns than ever before.  Furthermore, the commercialization of
virtual machines raises questions of how to ensure quality of service in
an environment where everyone is in contention for the hardware.  In this
paper, we will examine some of the recent literature in this area and
seek to draw some conclusions about the problem space and the directions
of future work.

The rest of the paper is structured as follows: in Section \ref{sec:background}
we provide background information on the history of virtualization and the
sorts of techniques which are utilized today.  Section \ref{sec:summaries}
presents summaries of a foundational virtualization paper from the
5\textsuperscript{th} USENIX Symposium on Operating Systems Design
and Implementation, as well as three recent papers from the virtualization
session of the same Symposium in 2010.  Section \ref{sec:relns} attempts
to analyze and classify these papers in terms of how they
relate to each other and how they approach the larger problem space.
Finally, in Section \ref{sec:conclusion} we conclude by discussing our
thoughts and opinions on the problem space as a whole, and postulate
about how virtualization may evolve in the future.

@@

Virtualization got it's start back in the 1960s, when developers
of large time-sharing systems sought ways to give each user the illusion that
they had sole access to a private, virtual computer, despite the fact that
many users were in fact sharing computing resources.  Once private physical
computers became available, popular, and cheap in the 80s, virtualization
took a backseat to other areas of research.  However, many of the ideas
put forth in these early days lived on in work involving virtual memory
and operating system protection paradigms.

Server virtualization arrived on the scene about a decade ago, with VMWare
appearing in 1999 and Xen in 2003.  Since then, research and commercial interest
in virtualization has sparked significantly, especially in conjunction
with the rise of Cloud computing and Infrastructure-as-a-service.

In a virtualized context, we have many \emph{virtual machines}(VM) sharing the same
physical hardware.  In a \emph{fully virtualized} system, each VM has the illusion
that it alone has sole access to the hardware, although in a \emph{paravirtualized}
system VMs may be modified to be aware of the fact that they are virtualized.  This
sharing of resources is possible because access to the hardware is regulated by
a special piece of software, the \emph{virtual machine monitor} or \emph{hypervisor}.
In many ways, the hypervisor plays the role of the kernel in a traditional operating
system - it contains the privileged software that can directly access the hardware.

The hypervisor (and potentially other administrative software) runs in the \emph{host}
domain (in Xen, this is called \emph{dom0}), while other OSes running in VMs are 
referred to as \emph{guests} (in Xen, all guests run in \emph{domU}).  It is
crucial that all guests are isolated from one another, so that the actions of one guest
cannot adversely affect the performance of any other.  This is a difficult property
to guarantee in a system where all guests are in contention for limited hardware resources.

In the early days of virtualization, there was little-to-no hardware support,
and so the multiplexing of VMs onto the physical hardware was accomplished by
\emph{emulating} hardware devices in the software for each VM.  This had the unfortunate
drawback of being extremely slow and inefficient, because the hardware was emulated by
the hypervisor, and so most device accesses by a guest VM resulted in multiple traps
into the hypervisor (and therefore multiple crossings of protection boundaries).
In the past decade or so, chipsets have become more virtualization-friendly,
and most architectures (notably the x86) now offer hardware-level support for
virtualization, along with a set of privileged instructions for making use of it.
This has allowed hardware virtualization to become fast enough to be viable,
so that virtualized systems are now reasonably competitive with their native
counterparts.

@@

Three years after VMWare's appearance, Denali \cite{ref:denali} sought to provide a virtualization
solution design with two goals in mind, scalability and security. Denali's authors
envisioned a separation between the management of physical hardware and the management
of software services, transforming how internet services could be deployed. Such a system 
would allow untrusted software to run on third party infrastructure, support the distribution 
of dynamic content in delivery systems, or provide virtual hosting infrastructure.

In order to meet this vision, the Denali project introduced \emph{paravirtualization} by providing
a simplified and pared-down emulation of the hardware for VMs to interface with.  Denali also
adopted complete isolation of guest VMs, and made VMs extremely lightweight, such that the system can
support over 10,000 VMs on commodity machines. Denali diverges from many 
related projects by sacrificing backwards compatibility for its primary objectives,
but its designers argue that this design tradeoff is well worth the benefits.

At the high level, Denali provides an isolation kernel (a small kernel operating system) running
on top of x86 hardware. The kernel is designed to host many untrusted applications which are isolated
and require little-to-no data sharing. Denali multiplexes between the guest operating systems and isolates them
using private virtualized namespaces. Some special privileges, like dowloading a remote VM or loading initial
disk images, are handled by a supervisor host VM.

The isolation kernel exports a simpler interface to guest VMs than would be required by a
traditional operating system.  Since the x86 ISA includes instructions whose behavior differs
in user and kernel mode, these are excluded from Denali's instruction set,
while two additional virtual instructions are added.
First, \verb+idle-with-timeout+ lets VMs halt their execution
for a bounded amount of physical time, or until an interrupt arrives for the VM,
which allows the system to avoid needlessly scheduling sleeping VMs.
Second, Denali provides a virtual instruction which a VM can use to self-terminate.

Denali's virtual architecture and implementation adhere to a separation between mechanism and policy 
so changes can be easily made. CPU virtualization holds a policy split into two elements, 
a \emph{gatekeeper policy} to choose the subset of active VMs to admit into the system, and 
a \emph{scheduler policy} to switch between active VMs. For simplicity, the authors use FIFO
and round robin in the descirbed implementation, although other scheduling policies could
be implemented.  The system provides a non-traditional interrupt model, where virtual interrupts
for a VM are batched and delivered asynchronously when the VM is next scheduled.

Denali's memory virtualization aims to keep a VM's working set in memory, so physical memory 
is redistributed from inactive to active VMs, and swap regions for each VM are used to page its memory.
The isolation kernel itself is kept pinned in memory at all times.

Denali also provides some limited emulated hardware.  In particular, a virtual ethernet NIC is 
provided for each VM, with queues for incoming and outgoing packets.
Incoming packets for each VM are accumulated in the appropriate queues and are later processed once the
corresponding VM is given focus by the scheduler.  Outgoing packets are processed in a
round-robin manner from outgoing queues. Denali's TCP/IP stack runs at the user level, which
has its drawbacks, since all packets have to cross the user/kernel boundary. However, the paper contains
a packet dispatch evaluation, which suggests that cost of network virtualization is low and that the the
driver itself is responsible for the majority of the costs of packet processing.

The Denali paper provides a comprehensive evaluation of the isolation kernel running
the Ilwaco operating system, with basic metrics like switching times between VMs, swap disk
microbenchmarks, and packet dispatch latency, and TCP and HTTP throughput. Some of the most
interesting evaluations were conducted to examine the scalability and performance of Denali.
The authors found that batched interrupts provided a performance gain of 30\% when the number
of VMs exceeded 800, and that using idle-with-timeout in Denali lead to higher performance as
the number of VMs increased.

However, Denali is certainly not free from bottlenecks, and the authors discuss two major challenges to
Denali's performance and scalability. First, Denali tries to reduce the amount of per-VM kernel data
that it has to manage by no keeping information for inactive VMs, but it still requires about 81MB
of data for 10,000 virtual machines. The authors suggest that with the growing size of memory,
this bottleneck might be acceptable. Secondly, the paging behavior of the guest is crucial to overall performance,
and while Denali does its best to keep VMs' working sets in memory, a misbehaving guest can severly damage the
performance of the system.  The authors claim that this problem may be solvable with good memory management 
and garbage collection techniques, which could increase locality.

@@

Turtles is designed for the x86 architecture, which subscribes to a
single-level hardware support model.  This means that there are only two
levels of execution privilege.  Turtles runs its host hypervisor, which
we will call $L_0$, at the most privileged level of execution.  Now, suppose
we have a hypervisor running above $L_0$ at level $L_1$, and further suppose 
that hypervisor is running a guest OS at level $L_2$.  Everything running at $L_1$
and $L_2$ has to run at the same privilege level, and only $L_0$ is privileged enough
to interact with the hardware.  Essentially, levels $L_1$ through $L_n$ only
exist logically; in reality they are all multiplexed to be running directly
on top of $L_0$ (see Figure \ref{fig:turtles-multiplex}).  All traps generated at
$L_2$ go immediately to $L_0$, which may handle them itself, or forward them
back to $L_1$ for handling if hardware interaction is not required.

@@

There are two difficult problems here that turtles provides a unique solution for.
The first involves page tables for the MMU.  In a nested model, virtual addresses for $L_2$
will need to be translated three times (first to physical addresses at $L_2$, then to $L_1$
addresses, and finally to $L_0$ addresses)
before they are readable by the hardware.  Modern MMUs provide support for virtualization
in the form of two distinct translation tables to make this process fast, but this is not
sufficient for more than one level of virtualization.  Furthermore, providing these
translation tables by emulating them in software is slow, particularly on updates,
for reasons described previously.  The paper's novel solution here is to use
\emph{multidimensional page tables}.  Essentially, each level keeps its own
shadowed page tables in software, but they are read-only.  On a page fault in $L_2$,
$L_0$ catches the trap and forwards it to $L_1$.  If $L_1$ also faults, $L_0$ allows
it to update its shadow table with the $L_2$-to-$L_1$ translation, but then applies
its tranlation for $L_1$ addresses to update the hardware-level table with the correct
$L_2$-to-$L_0$ translation before returning.  In this way, the hardware table at $L_0$
contains all the shadow tables above it compressed into a single table, so that
any MMU accesses can look up the correct translations directly from the hardware
without having to go through cascading levels of shadow tables first.  Turtles
also employs ``huge pages'' and Virtual Processor ID tags on page table entries
to ensure that page tables remain small and fast, and to prevent excessive TLB
flushing on context-switches.

The second difficulty deals with device I/0.  Traditional approaches to this problem
involve either software emulation (which is slow), installing modified drivers in the
guest (which is not preferrable), or assigning devices to guests as they have need
of them (which is hard in the nested case).  Turtles uses something called
\emph{multi-level device assignment}, which is essentially an adaptation of the
multidimensional page table idea to the device I/O problem.  Modern chipsets
include a component called an IOMMU, which resides between the DMA and main memory
and performs translations just like the MMU does for the CPU.  By shadowing read-only
IOMMU translation tables at higher levels, and then compressing them into a single
hardware-level table, the DMA can avoid the same cascading translation issue that
would affect the MMU, thus making DMA I/O access very efficient.  For Memory-Mapped I/O
and Network Port I/O, address translation is not an issue, and so here techniques for
single-level device assignment are used without any efficiency penalty, although
this does raise questions of fairness.

The Turtles paper provides an extensive evaluation section, with thorough experimentation
comparing performance of applications running on the native OS, a single-level VM, and the
Turtles two-level nested VM.  The authors measure and discuss performance on microbenchmarks,
I/O intensive workloads, and workloads that are designed to fault to the kernel constantly.
The paper also analyzes the success of multidimensional paging and multi-level device
assignment by testing these techniques against more traditional ways of handling these
problems.  In general, the Turtles hypervisor far outperforms any other approach to
nested virtualization considered, and is able to minimize the overhead of a second
level of virtualization to as low as 6-8\%, which is rather impressive.  No experiments
were conducted involving more than two nested levels of virtualization, although analysis
seemed to indicate that most of the additional overhead incurred by adding a second level
comes from cache pollution and the overuse of privileged instructions where it was not
necessary.

@@

The mClock paper \cite{ref:mclock} is not targeted at nested virtualization,
but further explores
the issue of fair I/O device assignment raised in the Turtles paper.  This turns
out to be a very difficult problem even in the single-level case, particularly
when applied to virtualization in the Cloud, where networked devices are
often shared not just between VMs running on one host, but between many hosts
across the network.  In the interest of fairness, a good hypervisor will try to
share device throughput equally across its guests, but in the networked case, the
total amount of capacity to share may fluctuate due to traffic from other
networked clients.  Furthermore, many OSes run their own device access algorithms
above the hypervisor level to improve performance; for example, physical disk
accesses may be reordered by the kernel to take advantage of spatial locality
when moving the disk head.  These sorts of optimizations can make it difficult
for the hypervisor to multiplex and schedule I/O requests efficiently.

The mClock algorithm seeks to fairly schedule the device I/O requests of multiple guests,
while maintaining each guest's limit (maximum throughput cap), reservation
(minimum requirement), and weight (proportional share of capacity relative to
other guests).  The algorithm uses a ``novel, lightweight'' tagging scheme to
achieve this despite fluctuation in total capacity.  Previous attempts to solve
this problem were not able to achieve both of these goals simultaneously.

@@

The mClock algorithm (see Figure \ref{fig:mclock-alg}) works by assigning numerical tags to each
queued request.  Essentially, tags for VM $i$ with proportional weight $w_i$ will be spaced by $1/w_i$.
Requests will be executed in the order of their assigned tags, which ensures that
VM $i$ receieves his proportional share.  This approach requires a global clock
for tag assignment, so that idle clients remain synced with active ones and do not
amass ``idle credit'', thus gaining preferential treatment.  The tag for VM $i$ and
request $j$ is set as follows:
\[
	Tag(i, j) = max(curr~time, Tag(i, j-1) + 1/w_i)
\]
After the tagging, the request with the lowest tag will be scheduled.

@@

The paper describes some further tweaks to the algorithm, such as allowing
a limited amount of idle credit to be saved up by idle VMs, or breaking ``large''
requests into a series of smaller requests to account for request latency
in the scheduler.  There's also a modifed dmClock algorithm for use with
a distributed storage system, which multiplexes VM requests fairly across multiple
storage servers.  Finally, there's a heuristic for setting reservations for
all VMs by default, so that no VM will be starved.

@@

The algorithm is successful at delivering on its guarantees despite contention,
which is important for ensuring VM isolation, and also has implications in the Cloud,
where the ability to meet quality-of-service guarantees has a high impact on
financial success.

@@

The ``Virtualize Everything but Time'' paper \cite{ref:time} notes that accurate
timekeeping has become very important in today's computing systems, with
applications to everything from systems measurement and distributed database
consistency to high-speed trading and billing for Cloud services.  However, 
timekeeping in a VM context is very difficult.  Historically, timekeeping
is accomplished using software clocks based on hardware oscillators, which
are kept synchronized to an online
reference clock.  However, the synchronization algorithms used rely on
computing delays between timestamped packets, and the delay computation
turns out to be extremely unstable in a virtualized context, due mostly
to added latency during packet processing.

@@

The focus of the paper is more on timekeeping synchronization algorithms.
The most popular contender in this arena is the \emph{Network Time Protocol} (NTP).
NTP is a feedback algorithm, which works by collecting and timestamping a series
of packets from a networked reference clock, computing clock error using
the reference timestamp-local timestamp pairings as data points, 
and updating the system clock accordingly.  Xen Clocksource makes use of NTP,
although it must be adapted to fit the virtualized context.  There are two
possibilities for this:
\begin{itemize}
\item in Dependent Clocksource, \emph{dom0} runs an NTP
clock, and all guests running in \emph{domU} sync periodically to this NTP
clock and use Clocksource to interpolate between syncs.  
\item under the Independent Clocksource paradigm, each \emph{domU} guest runs their own
NTP clock and send their own reference packets.
\end{itemize}
The paper conducts evaluation of both paradigms, and argues
that neither is well-suited to the virtualized context.  Dependent Clocksource
exhibits cyclical sawtoothed drift because the Clocksource algorithm in
\emph{domU} does not have direct access to the TSC, and so cannot adjust
the NTP clock data intelligently.  Independent Clocksource is extremely
inefficient, but even worse, additional latencies incurred by contention for
network I/O tend to push the Clocksource feedback algorithm into instability,
so that large error can accumulate even across NTP synchronizations.

@@

As we surveyed this series of papers, we noticed that the literature
in this area seems to neatly classify itself by \emph{motivation}.
Many papers on virtualization come from the realm of academia, and these
are mostly characterized by a focus on clean abstractions and solutions
with good theoretical properties, such as fairness and strong isolation.
Many more papers, however, originate from corporate research at places
like Microsoft and HP Labs.  These papers seem much more interested
in providing fast, scalable solutions using commodity hardware and
software.  In our papers, we found a series of tensions which seemed
to be expressing this fundamental question of motivation.  We will
now turn our attention to these tensions, keeping in mind that the
Turtles and mClock papers share corporate authorship, while the
Denali, Xen, and ``Virtualize Everything but Time'' papers originate
in the realm of academia.

@@

The most obvious tension which surfaced in the literature was a question
of priority between performance and fairness.  Generally speaking, there is a trend
in corporate papers to be heavily concerned with speed and efficiency, while
academically-minded papers seem more focused on strong properties of isolation
and guarantees of fairness between VMs.  This should not be particularly surprising,
given the authorship; more efficient virtualization seems a logical step toward maximizing
profit, which has a certain appeal in the corporate sphere.  It is worth noting, however,
that in the more recent literature corporate papers seem to be taking note of the
importance of fairness; we will discuss this further in Section \ref{sec:relns/scale}.

Our papers are clearly divided by this metric:  Turtles is all about efficiency
and performance, often at the cost of fairness.  The problem seems
especially hard here when phrased in the nested context.  It seems perfectly logical that if
a guest OS running in $L_2$ should only receive a fraction $\alpha$ of the resources allocated
to its host in $L_1$, which in turn should only receive a percentage $\beta$ of the machine's
total resources, then these two fractions should compound and our guest should end up with
$\alpha\beta$ of the total capacity available.  However, in reality Turtles multiplexes
all guests at all levels onto $L_0$, so in many ways the guest OS and its hypervisor will
be treated equally, rather than as parent and child.  Multi-level device assignment
for Port I/O and Memory-mapped I/O is particularly agnostic to fairness concerns, because
it works exactly the same way as single-level device assignment, and so doesn't care whether
a guest runs at $L_1$ or $L_{100}$ - all guests get equal treatment.  Furthermore, no
experimentation at all is done to analyze the effects of this model on fairness;
the authors even make the statement that this is a secondary concern,
and so they are content with a sort of best-effort.

On the other hand, Denali and mClock are all about prioritizing fairness.  The foundational
goal of the Denali isolation kernel is to ensure strong isolation between clients, so that
the behavior of any particular VM cannot in any way affect another.  And the whole point of
the mClock scheduling algorithm is to ensure that limits, reservations, and weights
are met fairly between VMs contending for a resource.  It should be noted that neither
of these papers particularly eschew performance as a worthy goal; rather, the view
here seems more to be that fast performance for a particular VM doesn't really
matter much if a malicious (or even just selfish) neighbor can somehow choke that VM
out and make it impossible to make progress.

@@

The split here is pretty clear as well: Denali, Xen, and Time all consider
paravirtualization to be beneficial and even necessary in the right situations.
In Denali, the isolation kernel presents a modified and simplified hardware abstraction
to its clients, which cleverly allows for some big wins in efficiency and performance
as a side effect.  Xen claims that paravirtualization is necessary to ensure fairness
and prevent starvation in process and I/O scheduling; one could argue, however, that
this is no longer a concern due to the advent of newer virtualization-aware hardware
(see Section \ref{sec:relns/hard-emul}).  The Time paper goes a step further and claims that
even with the new hardware, making the guest OS aware of the virtualized context it
runs in is necessary to ensure the correctness of certain time-dependent applications,
such as systems measurement, high-speed trading, and even video gaming.

In contrast, the Turtles project argues for full virtualization and unmodified guest
OSes and hypervisors.  In the paper, the argument for this goes back to the
motivating goal of performance, though it was unclear to us why paravirtualization
would have any negative impact here.  But it seems also that for any sort of large-scale
commercial deployment of server virtualization technology, the ability to use commodity
guests would be crucial in order to make adoption easy, and we consider this to also
be a likely factor influencing the design choice.
